---
title: "Methods for Handling Multiple Outcomes in Health Data"
subtitle: 'SECOMBIT Study'
author: "Sivakamakshi Muthu Kumarasamy, Shivadarshini Sreekanth"
output: 
  rmdformats::readthedown:
    self_contained: true
    highlight: "default"
    thumbnails: false
    fig_width: 7
    fig_height: 5
    toc_depth: 3

---
#libraries
```{r setup, message=FALSE, warning=FALSE}


library(dplyr)
library(ggplot2)
library(readxl)
library(survival)
library(survminer)
library(tableone)
library(mice)
library(plotly)
library(table1)
library(tidyr)
library(readr)
library(readxl)
library(dplyr)
library(table1)
library(survival)
library(survminer)
library(WinRatio) 
library(BuyseTest)
library(msm)
```

```{r}
SECOMBIT <- read.csv("Secombit data.csv", check.names = FALSE)

# Clean the data to improve readability
clean_df <- SECOMBIT %>%
  mutate(
    
    # ID
    id = ordine, 
    
    # Overall Survival
    os_time   = OS,
    os_status = status,
    
    # Progression-Free Survival
    pfs_time   = `PFS TOTAL`,
    pfs_status = `Progr total`,
    
    # Treatment arm (A as baseline)
    Arm = factor(ARM, levels = c("A", "B", "C")),
    
    # Number of sites (1–2 as baseline)
    Sites = factor(sites, levels = c("1-2", ">=3"),
                   labels = c("1 - 2", ">= 3")),
    
    # LDH (Normal as baseline)
    LDH = factor(ULN_LDH, levels = c("normal", "elevated"),
                 labels = c("Normal", "Elevated")),
    
    # TMB (<10 as baseline)
    TMB = factor(TMB, levels = c("<10", ">=10"),
                 labels = c("< 10", ">= 10")),
    
    # JAK (Wild Type as baseline)
    JAK = factor(JAK, levels = c("wt", "mut"),
                 labels = c("Wild Type (Normal)", "Mutated")),
    
  ) %>%
  select(id, Arm, Sites, LDH, TMB, JAK, pfs_time, pfs_status, os_time, os_status)


```
```{r}
names(clean_df)
```
## Numeric summary

```{r}

numeric_summary <- SECOMBIT %>%
summarise(
mean_OS    = mean(OS, na.rm = TRUE),
sd_OS      = sd(OS, na.rm = TRUE),
median_OS  = median(OS, na.rm = TRUE),
min_OS     = min(OS, na.rm = TRUE),
max_OS     = max(OS, na.rm = TRUE),
mean_PFS   = mean(`PFS TOTAL`, na.rm = TRUE),
sd_PFS     = sd(`PFS TOTAL`, na.rm = TRUE),
median_PFS = median(`PFS TOTAL`, na.rm = TRUE),
min_PFS    = min(`PFS TOTAL`, na.rm = TRUE),
max_PFS    = max(`PFS TOTAL`, na.rm =TRUE)
)

numeric_summary
```
## Summary statistics for categorical variables

```{r}

label(clean_df$Sites) <- "Sites"
label(clean_df$LDH)   <- "LDH"
label(clean_df$TMB)   <- "TMB"
label(clean_df$JAK)   <- "JAK"

table1(~ Sites + LDH + TMB + JAK | Arm,
       data = clean_df,
       overall = FALSE)


table_cat <- table1(~ LDH | Arm, data = clean_df)
table_cat

mortality_plot <- ggplot(clean_df, aes(x = Arm, fill = as.factor(LDH)))+
geom_bar(position = "dodge")+
labs(
title = "LDH levels by ARM",
x = "ARM",
y = "Levels",
fill = "LDH"
) +
scale_fill_manual(values = c("pink", "black", "yellow"),
labels = c("Missing","Normal", "Elevated")) +
theme_minimal()

ggplotly(mortality_plot)
```
## ARM-wise summary

```{r}
arm_summary <- clean_df %>%
group_by(Arm) %>%
summarise(
n = n(),
mean_OS = mean(os_status, na.rm = TRUE),
sd_OS = sd(os_status, na.rm = TRUE),
mean_PFS = mean(pfs_status, na.rm = TRUE),
sd_PFS = sd(pfs_status, na.rm = TRUE),
.groups = "drop"
)

arm_summary
```
```{r}
event_count <- clean_df %>%
  group_by(Arm)  %>%
  summarise(
    OS_Events = sum(os_status == 1, na.rm = TRUE),
    PFS_Events = sum(pfs_status == 1, na.rm = TRUE),
    .groups = "drop"
  )
event_count

event_long <- event_count %>%
  pivot_longer(
    cols = c(OS_Events, PFS_Events),
    names_to = "Endpoint",
    values_to = "Events"
  )

plot_events <- ggplot(event_long, aes(x = Arm, y = Events, fill = Endpoint)) +
  geom_col (position = position_dodge(width = 0.7))+
  labs(
    title = "OS and PFS : Event Vs Arms",
    x = "ARM",
    y = "No. of events",
    fill = "Endpoint",
  ) +
scale_fill_manual(values = c("OS_Events"  = "red", 
                             "PFS_Events" = "blue"))+
theme_minimal()

ggplotly(plot_events)


```
## Baseline variables

```{r}
vars <- c("Sites", "LDH", "TMB", "JAK", "os_status", "pfs_status")
factorVars <- c("Sites", "LDH", "TMB", "JAK")

tab1 <- CreateTableOne(
vars = vars,
strata = "Arm",
data = clean_df,
factorVars = factorVars
)

print(tab1, showAllLevels = TRUE, quote = FALSE, noSpaces = TRUE)
```
## **Composite Outcomes**

```{r}
clean_df$comp_status = ifelse(clean_df$os_status == 1 | clean_df$pfs_status == 1, 1, 0)
clean_df$comp_time = pmin(clean_df$os_time, clean_df$pfs_time, na.rm = TRUE)
 
table1(~comp_status + comp_time | Arm, data = clean_df)

```
## Binary outcome
```{r}

clean_df <- clean_df %>%
  mutate(
    binary_os = os_status,
    binary_pfs = pfs_status,
    binary_comp = comp_status
  ) %>%
  mutate(across(c(binary_os, binary_pfs, binary_comp), ~factor (.x, labels = c("No Event", "Event"))))


table(clean_df$binary_os, clean_df$Arm)
table(clean_df$binary_pfs, clean_df$Arm)
table(clean_df$binary_comp, clean_df$Arm)
```

## Logistic Regression
```{r}
log_os <- glm(binary_os ~ Arm + Sites + LDH + TMB + JAK, family = binomial, data = clean_df)

log_pfs <- glm(binary_pfs ~ Arm + Sites + LDH + TMB + JAK, family = binomial, data = clean_df)

log_comp <- glm(binary_comp ~ Arm + Sites + LDH + TMB + JAK, family = binomial, data = clean_df)

tidy_logistic <- function(fit, title){
  broom::tidy(fit, exponentiate = TRUE, conf.int = TRUE) |>
    mutate(
      OR_CI = sprintf("%.2f (%.2f–%.2f)", estimate, conf.low, conf.high),
      p = signif(p.value, 3)
    ) |>
    select(term, OR_CI, p) |>
    rename(Variable = term, `OR (95% CI)` = OR_CI, `p-value` = p) |>
    (function(df){
      cat(paste0("<b>", title, "</b>"))
      knitr::kable(df, format = "html", escape = FALSE) |>
        kableExtra::kable_styling(full_width = TRUE, font_size = 18)
    })()
}

tidy_logistic(log_os, "Logistic — OS Event")
cat("<br>")
tidy_logistic(log_pfs, "Logistic — PFS Event") 
cat("<br>")
tidy_logistic(log_comp, "Logistic — Composite Event")
```

## Logistic Regression and COX for ARM

```{r}
log_os_arm <- glm(binary_os ~ Arm, family = binomial, data = clean_df)

log_pfs_arm <- glm(binary_pfs ~ Arm, family = binomial, data = clean_df)

log_comp_arm <- glm(binary_comp ~ Arm, family = binomial, data = clean_df)

tidy_logistic <- function(fit, title){
  broom::tidy(fit, exponentiate = TRUE, conf.int = TRUE) |>
    mutate(
      OR_CI = sprintf("%.2f (%.2f–%.2f)", estimate, conf.low, conf.high),
      p = signif(p.value, 3)
    ) |>
    select(term, OR_CI, p) |>
    rename(Variable = term, `OR (95% CI)` = OR_CI, `p-value` = p) |>
    (function(df){
      cat(paste0("<b>", title, "</b>"))
      knitr::kable(df, format = "html", escape = FALSE) |>
        kableExtra::kable_styling(full_width = TRUE, font_size = 18)
    })()
}

tidy_logistic(log_os_arm, "Logistic — OS Event by ARM")
cat("<br>")
tidy_logistic(log_pfs_arm, "Logistic — PFS Event by ARM") 
cat("<br>")
tidy_logistic(log_comp_arm, "Logistic — Composite Event by ARM")
```
```{r}
cox_os_arm  <- coxph(Surv(os_time, os_status) ~ Arm,
                 data = clean_df, ties = "efron")
cox_pfs_arm <- coxph(Surv(pfs_time, pfs_status) ~ Arm,
                 data = clean_df, ties = "efron")

cox_comp_arm <- coxph(Surv(comp_time, comp_status) ~ Arm,
                  data = clean_df, ties = "efron")

tidy_cox <- function(fit, title){
  broom::tidy(fit, exponentiate = TRUE, conf.int = TRUE) |>
    mutate(
      HR_CI = sprintf("%.2f (%.2f–%.2f)", estimate, conf.low, conf.high),
      p = signif(p.value, 3)
    ) |>
    select(term, HR_CI, p) |>
    rename(Variable = term, `HR (95% CI)` = HR_CI, `p-value` = p) |>
    (\(df){
      cat(paste0("<b>", title, "</b>"))
      knitr::kable(df, format="html", escape=TRUE) |>
        kableExtra::kable_styling(full_width=TRUE, font_size=18)
    })()
}

tidy_cox(cox_os_arm,  "Cox model — Overall Survival (OS) by ARM")
cat("<br>")
tidy_cox(cox_pfs_arm, "Cox model — Progression-Free Survival (PFS) by ARM")
cat("<br>")
tidy_cox(cox_comp_arm, "Cox model — Composite Outcome by ARM")


```

## Univariate Analysis

## KM Plot - OS and PFS

```{r}
par(mar = c(3, 3, 1.1, 1))
fit_os <- survfit(Surv(os_time, os_status) ~ Arm, data = clean_df)

os_km <- ggsurvplot(
  fit_os, data = clean_df,
  xlab = "Time (months)", ylab = "Survival probability",
  title = "Overall Survival (OS)",
  risk.table = FALSE, pval = TRUE, conf.int = TRUE, pval.coord = c(10,0.5), 
)

#os_km$plot <- os_km$plot + geom_vline(xintercept = 2, linetype = "dashed") +
  #annotate("text", x = 2, y = 0.2, label = "2 months", angle = 90, vjust = -1)

os_km
```


```{r}
par(mar = c(3, 3, 1.1, 1))
fit_pfs <- survfit(Surv(pfs_time, pfs_status) ~ Arm, data = clean_df)

pfs_km <- ggsurvplot(
  fit_pfs, data = clean_df,
  xlab = "Time (months)", ylab = "Survival probability",
  title = "Progression-Free Survival (PFS)",
 risk.table = FALSE, pval = TRUE, conf.int = TRUE, pval.coord = c(10,0.2)
) 

#pfs_km$plot <- pfs_km$plot + geom_vline(xintercept = 2, linetype = "dashed")+
  #annotate("text", x = 2, y = 0.2, label = "2 months", angle = 90, vjust = -1)

pfs_km
```
## KM plot for composite outcome

```{r}
par(mar = c(3, 3, 1.1, 1))
fit_comp <- survfit(Surv(comp_time, comp_status) ~ Arm, data = clean_df)

comp_km <- ggsurvplot(
  fit_comp, data = clean_df,
  xlab = "Time (months)", ylab = "Survival probability",
  title = "Composite Outcome",
  risk.table = FALSE, pval = TRUE, conf.int = TRUE, 
  legend.title = "Arm", legend.labs = levels(clean_df$Arm),pval.coord = c(10,0.2)
)
#comp_km$plot <- comp_km$plot + geom_vline(xintercept = 2, linetype = "dashed")+
  #annotate("text", x = 2, y = 0.2, label = "2 months", angle = 90, vjust = -1)

comp_km
```


## KM curves until 2 months

```{r, warning=FALSE}
os_km_2 <- ggsurvplot(
  fit_os, data = clean_df,
  xlab = "Time (months)", ylab = "Survival probability",
  title = "Overall Survival (OS)",
  risk.table = FALSE, pval = TRUE, pval.coord = c(10,0.5), xlim = c(0,4), ylim = c(0.9,1.0),break.time.by = 1
)

os_km_2

```

```{r, warning=FALSE}
pfs_km_2 <- ggsurvplot(
  fit_pfs, data = clean_df,
  xlab = "Time (months)", ylab = "Survival probability",
  title = "Progression-Free Survival (PFS)",
 risk.table = FALSE, pval = TRUE, pval.coord = c(10,0.2), xlim = c(0,4), ylim = c(0.9,1.0), break.time.by = 1
) 

pfs_km_2

```
```{r, warning=FALSE}
comp_km_2 <- ggsurvplot(
  fit_comp, data = clean_df,
  xlab = "Time (months)", ylab = "Survival probability",
  title = "Composite Outcome",
  risk.table = FALSE, pval = TRUE,
  legend.title = "Arm", legend.labs = levels(clean_df$Arm),pval.coord = c(10,0.2), xlim = c(0,4), ylim = c(0.9,1.0), break.by.time = 1
)

comp_km_2

```

## LogRank test - OS and PFS

```{r}


logrank_os  <- survdiff(Surv(os_time, os_status) ~ Arm, data = clean_df)
logrank_pfs <- survdiff(Surv(pfs_time, pfs_status) ~ Arm, data = clean_df)


p_os  <- 1 - pchisq(logrank_os$chisq,  df = length(logrank_os$n)  - 1)
p_pfs <- 1 - pchisq(logrank_pfs$chisq, df = length(logrank_pfs$n) - 1)

logrank_tab <- data.frame(
  Endpoint = c("OS", "PFS"),
  `Chi-square` = round(c(logrank_os$chisq, logrank_pfs$chisq), 2),
  df = c(length(logrank_os$n)-1, length(logrank_pfs$n)-1),
  `p-value` = signif(c(p_os, p_pfs), 3)
)

knitr::kable(logrank_tab, format = "html", align = "lccc") |>
  kableExtra::kable_styling(full_width = TRUE, font_size = 16)

```

**Inference - Survival curves and Cox Log-Rank tests**

Survival curves and log‑rank tests
The Kaplan–Meier curves for OS appear relatively close, and the log‑rank test p‑value around 0.25 suggests no statistically significant difference in overall survival across arms at this stage.

The PFS curves separate more clearly, with a log‑rank p‑value near 0.02, indicating evidence that at least one arm differs in progression‑free survival, consistent with SECOMBIT reports that sequencing can affect progression endpoints.

# LogRank test - Composite Outcome

```{r}
logrank_comp  <- survdiff(Surv(comp_time, comp_status) ~ Arm, data = clean_df)

p_comp  <- 1 - pchisq(logrank_comp$chisq,  df = length(logrank_comp$n)  - 1)


logrank_tab <- data.frame(
  Endpoint = "Composite Outcome",
  `Chi-square` = round((logrank_comp$chisq), 2),
  df = length(logrank_comp$n)-1,
  `p-value` = signif(p_comp, 3)
)

knitr::kable(logrank_tab, format = "html", align = "lccc") |>
  kableExtra::kable_styling(full_width = TRUE, font_size = 16)

```

## Cox proportional Hazards model - OS and PFS


```{r}
library(broom)

cox_os  <- coxph(Surv(os_time, os_status) ~ Arm + Sites + LDH + TMB + JAK,
                 data = clean_df, ties = "efron")
cox_pfs <- coxph(Surv(pfs_time, pfs_status) ~ Arm + Sites + LDH + TMB + JAK,
                 data = clean_df, ties = "efron")

tidy_cox <- function(fit, title){
  broom::tidy(fit, exponentiate = TRUE, conf.int = TRUE) |>
    mutate(
      HR_CI = sprintf("%.2f (%.2f–%.2f)", estimate, conf.low, conf.high),
      p = signif(p.value, 3)
    ) |>
    select(term, HR_CI, p) |>
    rename(Variable = term, `HR (95% CI)` = HR_CI, `p-value` = p) |>
    (\(df){
      cat(paste0("<b>", title, "</b>"))
      knitr::kable(df, format="html", escape=TRUE) |>
        kableExtra::kable_styling(full_width=TRUE, font_size=18)
    })()
}

tidy_cox(cox_os,  "Cox model — Overall Survival (OS)")
cat("<br>")
tidy_cox(cox_pfs, "Cox model — Progression-Free Survival (PFS)")


```
The multivariable Cox proportional hazards model showed no statistically significant associations between the examined covariates and survival outcome (both OS and PFS). All variables had hazard ratios with 95% confidence intervals that included 1.0 and p-values exceeding 0.05.

**COX Model for OS:**

In the Cox model for OS, the hazard ratios for Arm B and Arm C versus Arm A are close to 1 with wide confidence intervals and non‑significant p‑values, suggesting no clear survival advantage yet for any single sequencing strategy.

Some baseline factors, such as number of metastatic sites and LDH elevation, typically show increased mortality risk in melanoma trials, so if their HRs are above 1 with low p‑values, this aligns with prior evidence that greater tumour burden and elevated LDH are adverse prognostic markers.

**COX Model for PFS:**

The Cox model results for progression‑free survival show no statistically significant effects for treatment arms or most covariates, though JAK mutation and Arm B trends toward better outcomes.

JAK mutated has the strongest signal with HR 0.41 (95% CI: 0.16–1.03, p=0.057), nearly halving progression risk and just missing significance; this matches SECOMBIT biomarker findings where JAK‑mutated patients had unexpectedly good PFS and brain metastasis‑free survival.

Sites ≥3 (HR 1.60, 95% CI: 0.80–3.20, p=0.187) and elevated LDH (HR 1.14, 95% CI: 0.56–2.34, p=0.715) show increased risk trends, reflecting tumour burden as an adverse factor, though wide CIs suggest limited power.

TMB ≥10 has HR 0.85 (95% CI: 0.40–1.78, p=0.666), a neutral protective trend, in line with high TMB benefiting immunotherapy‑containing arms.

Covariates such as elevated LDH, high TMB, or multiple sites, when associated with HRs above 1, reflect faster progression, again consistent with external evidence that high tumour burden and certain biomarker profiles predict shorter progression‑free survival.

## Cox proportional Hazards model - Composite Outcome

```{r}
library(broom)

cox_comp  <- coxph(Surv(comp_time, comp_status) ~ Arm + Sites + LDH + TMB + JAK,
                 data = clean_df, ties = "efron")

tidy_cox_comp <- function(fit, title){
  broom::tidy(fit, exponentiate = TRUE, conf.int = TRUE) |>
    mutate(
      HR_CI_comp = sprintf("%.2f (%.2f–%.2f)", estimate, conf.low, conf.high),
      p = signif(p.value, 3)
    ) |>
    select(term, HR_CI_comp, p) |>
    rename(Variable = term, `HR (95% CI)` = HR_CI_comp, `p-value` = p) |>
    (\(df){
      cat(paste0("<b>", title, "</b>"))
      knitr::kable(df, format="html", escape=TRUE) |>
        kableExtra::kable_styling(full_width=TRUE, font_size=18)
    })()
}

tidy_cox_comp(cox_comp,  "Cox model — Composite Outcome")
cat("<br>")
```

### PH Test

```{r ph-global, results="asis", echo=FALSE}
ph_os  <- cox.zph(cox_os)
ph_pfs <- cox.zph(cox_pfs)


ph_tab <- data.frame(
  Endpoint = c("OS", "PFS"),
  `PH test p-value` = signif(c(ph_os$table["GLOBAL","p"], ph_pfs$table["GLOBAL","p"]), 3)
)

knitr::kable(ph_tab, format="html") |>
  kableExtra::kable_styling(full_width=TRUE, font_size=18)
```

**Summary**

The PH test p-values indicate moderate violations of the proportional hazards assumption in the OS model and a more serious violation in the PFS model.

**OS model (p = 0.0581)**

This borderline p-value suggests the proportional hazards assumption holds approximately for overall survival, but there's weak evidence against it.

Hazard ratios from the Cox model remain reasonably interpretable, though caution is warranted; the curves may cross or diverge over time.


**PFS model (p = 0.0119)**

The low p-value clearly rejects proportional hazards for progression-free survival, meaning treatment effects likely change over follow-up time.

This justifies moving beyond standard Cox to time-varying effects, stratified models, or multi-state approaches.


```{r}
ph_comp  <- cox.zph(cox_comp)

ph_tab <- data.frame(
  Endpoint = "Composite outcome",
  `PH test p-value` = signif((ph_comp$table["GLOBAL","p"]),3)
)

knitr::kable(ph_tab, format="html") |>
  kableExtra::kable_styling(full_width=TRUE, font_size=18)

```

## Win-ratio

## B Vs A
```{r}
wr_ba <- winratio(id = "id",
                  trt = "Arm",
                  outcomes = list(
                    outc1 = c("os_status", "s", "os_time"),
                    outc2 = c("pfs_status", "s", "pfs_time")
                  ),
                  fu = "os_time",
                  data = clean_df %>% 
                    filter(Arm %in% c("A", "B"))%>%
                    filter(!is.na(os_time) & !is.na(os_status),
                           !is.na(pfs_time) & !is.na(pfs_status)) %>%
                    mutate(
      pfs_time = pmin(pfs_time, os_time),
      pfs_status = ifelse(os_status == 1 & pfs_time == os_time, 1, pfs_status)
    )
)
summary(wr_ba)
```
* This win-ratio analysis favours ARM B over ARM A with a clinically meaningful point estimate but lacks statistical significance.

* Out of 4761 paired comparisons between Arms A and B (with 69 patients in each arm), ARM B **wins** 2289 pairs while ARM A wins 1541, with a win-ratio of 1.49 and a p-value of 0.0997.

* Wins are more driven by OS (1733 Vs 1271) than PFS (556 Vs 270). This reflects ARM B's edge in both endpoints despite individual COX models with non-significant Hazard ratios.

* This aligns with SECOMBIT's longer-term data favoring Arms B & C.


## C Vs A
```{r}

wr_ca <- winratio(id = "id",
                  trt = "Arm",
                  outcomes = list(
                    outc1 = c("os_status", "s", "os_time"),
                    outc2 = c("pfs_status", "s", "pfs_time")
                  ),
                  fu = "os_time",
                  data = clean_df %>% filter(Arm != "B") %>%
                    filter(Arm %in% c("C", "A")) %>%
                    filter(!is.na(os_time), !is.na(os_status),
                           !is.na(pfs_time), !is.na(pfs_status)) %>%
                    mutate(
                      pfs_time = pmin(pfs_time, os_time),
                      pfs_status = ifelse(os_status == 1 & pfs_time == os_time, 1, pfs_status)
    )
)

summary(wr_ca)
```
The win-ratio analysis show Arm C has a statistically near-significant advantage over Arm A, with a superior performance across OS and PFS.

* Out of 4692 paired comparisons between Arms C and A (with 68 patients in Arm C and 69 in Arm A), ARM C wins 2362 pairs while ARM A wins only 1489, with a win-ratio of 1.59 and p-value is at 0.0503.

* Strong OS contribution (1850 Vs 1347) dominates, with PFS also favouring ARM C. 

* Few ties (841) than other analyses sharpen the signal.

* Both Arm C and A trend toward benefit over Arm A, with Arm C has nominal significance despite similar sample sizes.

## C Vs B
```{r}

wr_cb <- winratio(id = "id",
                  trt = "Arm",
                  outcomes = list(
                    outc1 = c("os_status", "s", "os_time"),
                    outc2 = c("pfs_status", "s", "pfs_time")
                  ),
                  fu = "os_time",
                  data = clean_df %>%
                    filter(Arm %in% c("C", "B")) %>%
                    filter(!is.na(os_time), !is.na(os_status),
                           !is.na(pfs_time), !is.na(pfs_status)) %>%
                    mutate(
                      pfs_time = pmin(pfs_time, os_time),
                      pfs_status = ifelse(os_status == 1 & pfs_time == os_time, 1, pfs_status)
    )
)

summary(wr_cb)
```
* The win-ratio analysis show no meaningful advantage for Arm C over Arm B, with a near-neutral point estimate and no statistical significance.

* Out of 4692 paired comparisons between Arms C and B (with 68 patients in Arm C and 69 in Arm B), ARM C wins 1647 pairs while ARM B wins 1561, with a win-ratio of 1.06 and p-value is at 0.8372.

* OS wins are balanced with PFS slightly favouring ARM C. 

* High ties and similar totals indicates equivalence.

* Both Arm B and C outperform A without any clear superiority between them.


## Adjusted p-values for multiple testing

```{r}

pvals <- c(
  "B_vs_A" = wr_ba$p.value,
  "C_vs_B" = wr_cb$p.value,
  "C_vs_A" = wr_ca$p.value
)

p.adjust(pvals, method = "bonferroni")

```
After multiple testing adjustment using Bonferroni, none of the win-ratio comparisions between arms shows statistical significance.

* **B Vs A:** Original p=0.0997 becomes 0.299 (non-significant)

* **C Vs B:** Original p=0.8372 becomes 1.000 (non-significant)

* **C Vs A:** Original p=0.0503 becomes 0.151 (non-significant)

* Bonferroni correction for 3 comparisons appropriately controls arm-wise error but reduces power with approximately 68-69 patients in each arm

* Unadjusted trends - Arm B and C both favor over Arm A but lacks confirmation.

## NET BENEFIT

# B Vs A

```{r}
nb_df_BA <- clean_df %>%
  filter(Arm %in% c("A","B")) %>%
  filter(!is.na(os_time) | !is.na(pfs_time)) %>%   # keep if at least one endpoint present
  mutate(
    trt = factor(Arm, levels = c("A","B"))
  )

nb_BA <- BuyseTest(
  trt ~ 
    tte(os_time, os_status, threshold = 0) +   # priority 1: OS
    tte(pfs_time, pfs_status, threshold = 0),  # priority 2: PFS
  data = nb_df_BA,
  hierarchical = TRUE
)

summary(nb_BA)

```
* Significant PFS contribution (p-value 0.031)

* Global results reflects 22% net benefit for B with OS alone contributing for 13% with 8.6% from PFS. 
 
# C Vs A

```{r}

nb_df_CA <- clean_df %>%
  filter(Arm %in% c("A","C")) %>%
  filter(!is.na(os_time) | !is.na(pfs_time)) %>%
  mutate(trt = factor(Arm, levels = c("A","C")))   # A control, C treatment

nb_CA <- BuyseTest(
  trt ~ 
    tte(os_time, os_status, threshold = 0) +
    tte(pfs_time, pfs_status, threshold = 0),
  data = nb_df_CA,
  hierarchical = TRUE
)

summary(nb_CA)

```
* Generalised pairwise comparisons for Arm C vs Arm A is similar to Arm B Vs Arm A.

* Global net benefit :  0.2015 - Arm C beats Arm A by overall 20.15% with 13.23% contributed by the OS. 

* Strong PFS signal noted (p=0.0499)

# C Vs B

```{r}
nb_df_CB <- clean_df %>%
  filter(Arm %in% c("B","C")) %>%
  filter(!is.na(os_time) | !is.na(pfs_time)) %>%
  mutate(trt = factor(Arm, levels = c("B","C")))   # B control, C treatment

nb_CB <- BuyseTest(
  trt ~ 
    tte(os_time, os_status, threshold = 0) +
    tte(pfs_time, pfs_status, threshold = 0),
  data = nb_df_CB,
  hierarchical = TRUE
)

summary(nb_CB)

```
* Generalised pairwise comparisons confirm equivalence between Arms C and B with no net benefit.

* OS contribution -1% net confirming long term survival, PFS -3.89% is clinically not significant.


## Multi-state model


State 1: Stable

State 2: Progression

State 3: Death

```{r}
states <- c("Stable", "Progression", "Death")

```

```{r}
#Define transition intensity matrix (Q)

Q <- matrix(0, nrow = 3, ncol = 3, dimnames = list(states, states))

#Other transitions remain 0 (impossible)

Q["Stable","Progression"] <- 1

#Q["Stable","Death"]       <- 1 though this is logically possible, it is not seen in this dataset

Q["Stable","Death"] <- 0
Q["Progression","Death"]  <- 1


```


```{r}
### Prepare data in msm long format


msm_df <- clean_df %>%
  filter(
    !is.na(pfs_time),
    !is.na(pfs_status),
    !is.na(os_time),
    !is.na(os_status)
  ) %>%
  mutate(
    Arm = factor(Arm, levels = c("A","B","C")),
    
    # set last follow-up time
    last_time = case_when(
      os_status == 1 ~ os_time,
      !is.na(os_time) | !is.na(pfs_time) ~ pmax(os_time, pfs_time, na.rm = TRUE),
      TRUE ~ 0
    )
  ) %>%
  rowwise() %>%
  do({
    id  <- .$id
    arm <- .$Arm
    sites <- .$Sites
    ldh <- .$LDH
    tmb <- .$TMB
    jak <- .$JAK
    
    rows <- list(
      # Everyone starts in Stable
      data.frame(id = id, time = 0, state = 1, Arm = arm, Sites = sites, LDH = ldh, TMB  = tmb, JAK = jak)
    )
    # Progression
   
    
    if (.$pfs_status == 1) {
      
      prog_time <- .$pfs_time
      
      rows <- c(rows, list(
        data.frame(id = id, time = prog_time, state = 2, Arm = arm, Sites = sites, LDH = ldh, TMB  = tmb, JAK = jak)
      ))
    }
    
    # Death
  
    if (.$os_status == 1) {
      
      death_time <- .$os_time
      
      # Ensure strict ordering if progression exists
      if (.$pfs_status == 1 && death_time <= .$pfs_time) {
        death_time <- .$pfs_time + 1e-5
      }
      
      rows <- c(rows, list(
        data.frame(id = id, time = death_time, state = 3, Arm = arm, Sites = sites, LDH = ldh, TMB  = tmb, JAK = jak)
      ))
      
    } else {
      
      # Censoring (no death)
      
      last_time <- .$os_time  # follow-up time
      
      last_state <- ifelse(.$pfs_status == 1, 2, 1)
      
      # Do NOT duplicate row if progression already at same time
      
      if (!(.$pfs_status == 1 && .$pfs_time == last_time)) {
        
        rows <- c(rows, list(
          data.frame(id = id, time = last_time, state = last_state, Arm = arm, Sites = sites, LDH = ldh, TMB  = tmb, JAK = jak)
        ))
      }
    }
    
    bind_rows(rows)
  }) %>%
  ungroup() %>%
  arrange(id, time)




```

## Multi-state  model

```{r}
# Arm as covariate affecting all transitions
msm_fit <- msm(
  state ~ time,
  subject = id,
  data = msm_df,
  qmatrix = Q,
  covariates = ~ Arm,
  exacttimes = TRUE
)

# Summary
summary(msm_fit)
```

**Interpretation:**

* Strong decline in number of patients in **Stable** over time.

* Death increases over time.

* Progression acts as the intermediate state between these two.

* Transition HR show that both experimental arms reduce the risk of Stable to Progression compared with Arm A. HR : 0.56, 95% CI - 0.35 - 0.88) for Arm B and 0.59 (95% CI 0.38 - 0.92) for Arm C, indicating statistically significant slowing of disease progression from the stable state.

* Neither Arm has a clear effect on Progression to Death transition. ARM B HR : 1.37 and Arm C HR : 1.02, both includes 1, so there is no evidence taht treatment changes post-progression mortality.

* Most patients transitions from Stable to Death, directly or via Progression but Arms B and C delays Progression from the Stable state without significantly altering the risk of death once Progression has occurred.

```{r}
# MSM with all covariates

msm_fit_all <- msm(
  state ~ time,
  subject = id,
  data = msm_df,
  qmatrix = Q,
  covariates = ~  Arm + Sites + LDH + TMB + JAK,
  exacttimes = TRUE
)

# Summary
summary(msm_fit_all)

```

## Transition probabilities

# Probabilities from Stable at time t

```{r}
pt <- pmatrix.msm(msm_fit, t = 12) 
pt

# Plot : Transition probabilities from Stable over time
times <- seq(0, 72, by = 1) 
probs <- sapply(times, function(t) pmatrix.msm(msm_fit, t = t)[1, ])
matplot(times, t(probs), type = "l", lty = 1, col = 1:3,
        xlab = "Time (months)", ylab = "Probability",
        main = "State occupation probabilities from Stable")
legend("topright", legend = states, col = 1:3, lty = 1)
```


```{r}
pt_A_12 <- pmatrix.msm(msm_fit, t = 12, covariates = list(Arm = "A"))
pt_B_12 <- pmatrix.msm(msm_fit, t = 12, covariates = list(Arm = "B"))
pt_C_12 <- pmatrix.msm(msm_fit, t = 12, covariates = list(Arm = "C"))

pt_A_12
pt_B_12
pt_C_12
```
**Arm A:**

* If starting Stable, 72.8% remain stable, 14.7% progress and 12.5% die.

* If starting Progression, 27.3% keeps progressing and 72.7% die.

**Arm B:**

* If starting Stable, 83.8% remain stable, 7.4% progress and 8.8% die.

* If starting Progression, 17% keeps progressing and 83% die.

**Arm C:**

* If starting Stable, 83% remain stable, 9.2% progress and 7.8% die.

* If starting Progression, 26.6% progress and 73.4% die.

**Conclusion:**

* Arm B and C prolong time in Stable state (83-84% Vs Arm A at 73%), with lower direct death risk from stable (8-9% Vs 12.5%).

* Post-progression trajectories are similar across arms, suggessting treatment delays progression onset without altering mortality once progressed aligning with hazard ratios for Stable to Progression (0.56-0.59).


```{r}
probs_A <- sapply(times, function(t)
  pmatrix.msm(msm_fit, t = t, covariates = list(Arm = "A"))[1, ])

probs_B <- sapply(times, function(t)
  pmatrix.msm(msm_fit, t = t, covariates = list(Arm = "B"))[1, ])

probs_C <- sapply(times, function(t)
  pmatrix.msm(msm_fit, t = t, covariates = list(Arm = "C"))[1, ])


```


## Plots 

# Probability of remaining Stable by Treatment

```{r}

plot(times, probs_A[1,], type = "l", lwd = 2,
     ylim = c(0,1),
     xlab = "Time (months)",
     ylab = "P(Stable)",
     main = "Probability of remaining Stable by Treatment")

lines(times, probs_B[1,], lwd = 2, lty = 2)
lines(times, probs_C[1,], lwd = 2, lty = 3)

legend("topright",
       legend = c("Arm A", "Arm B", "Arm C"),
       lty = c(1,2,3), lwd = 2)
```


# Cumulative probability of Death by Treatment

```{r}
plot(times, probs_A[3,], type = "l", lwd = 2,
     ylim = c(0,1),
     xlab = "Time (months)",
     ylab = "P(Death)",
     main = "Cumulative probability of Death by Treatment")

lines(times, probs_B[3,], lwd = 2, lty = 2)
lines(times, probs_C[3,], lwd = 2, lty = 3)

legend("topleft",
       legend = c("Arm A", "Arm B", "Arm C"),
       lty = c(1,2,3), lwd = 2)

```

# Probability of Death conditioned along Progression

```{r}
#Extracting probabilites of Progression and Death for each treatment arm

times <- seq(0, 72, by = 1) 
prob_death_A <- sapply(times, function(t)
  pmatrix.msm(msm_fit, t = t, covariates = list(Arm = "A"))[2, 3])

probs_death_B <- sapply(times, function(t)
  pmatrix.msm(msm_fit, t = t, covariates = list(Arm = "B"))[2, 3])

probs_death_C <- sapply(times, function(t)
  pmatrix.msm(msm_fit, t = t, covariates = list(Arm = "C"))[2, 3])


#Plot

plot(times, prob_death_A, type = "l",lwd = 2,
     ylim = c(0,1),
     xlab = "Time since Progression (months)",
     ylab = "P(Death | Progression)",
     main = "Probability of Death conditioned along Progression by Treatment")

lines(times, probs_death_B, lwd = 2, lty = 2)
lines(times, probs_death_C, lwd = 2, lty = 3)

legend("bottomright", legend = c("Arm A", "Arm B", "Arm C"), lty = c(1,2,3),lwd = 2)
```

# Long-term state occupancy

## Probailities across arms

```{r}

timeline <- seq(0,72, by = 2)
summary_df <- data.frame(
  Time = timeline,
  Stable_A = round(probs_A[1, match(timeline, times)],2),
  Stable_B = round(probs_B[1, match(timeline, times)],2),
  Stable_C = round(probs_C[1, match(timeline, times)],2),
  Death_A  = round(probs_A[3, match(timeline, times)],2),
  Death_B  = round(probs_B[3, match(timeline, times)],2),
  Death_C  = round(probs_C[3, match(timeline, times)],2)
)

summary_df
```

**Summary:**

* Arms B and C maintain higher probabilities of remaining stable over time compared to Arm A, with slower declines and lower death probabilities throughout follow-up.

* At 12 months, Stable probablities A < B < C, Death A > B > C.

* Arm B shows fastest initial stable retention (0.92) followed by C (0.91), with A at 0.85.

* By 60 months, Stable A(0.20) < C(0.39) < B (0.41); whereas Death A (0.73) > B & C (0.54).

* At the end of 66 months, Stable A (0.15) << B(0.35), C().36) while Death A(0.77) > B (0.58), C(0.62).

* Arm A consistently shows lowest stable probabilities and highest death probabilities at every time point.

* Arm B performs best long-term while maintaining highest stable state and lowest mortality.

* Arm C shows intermediate but superior to A outcomes throughout the time period.





