---
title: "SECOMBIT Study - Data exploration"
output:
  html_document: default
  word_document: default
date: "2025-11-24"
---

```{r setup, message=FALSE, warning=FALSE}
#libraries

library(dplyr)
library(ggplot2)
library(readxl)
library(survival)
library(survminer)
library(tableone)
library(mice)
library(plotly)
library(table1)
library(tidyr)
library(readr)
library(readxl)
library(dplyr)
library(table1)
library(survival)
library(survminer)
library(WinRatio) 
library(BuyseTest)
library(msm)
```

```{r}
SECOMBIT <- read.csv("Secombit data.csv", check.names = FALSE)

# Clean the data to improve readability
clean_df <- SECOMBIT %>%
  mutate(
    
    # ID
    id = ordine, 
    
    # Overall Survival
    os_time   = OS,
    os_status = status,
    
    # Progression-Free Survival
    pfs_time   = `PFS TOTAL`,
    pfs_status = `Progr total`,
    
    # Treatment arm (A as baseline)
    Arm = factor(ARM, levels = c("A", "B", "C")),
    
    # Number of sites (1–2 as baseline)
    Sites = factor(sites, levels = c("1-2", ">=3"),
                   labels = c("1 - 2", ">= 3")),
    
    # LDH (Normal as baseline)
    LDH = factor(ULN_LDH, levels = c("normal", "elevated"),
                 labels = c("Normal", "Elevated")),
    
    # TMB (<10 as baseline)
    TMB = factor(TMB, levels = c("<10", ">=10"),
                 labels = c("< 10", ">= 10")),
    
    # JAK (Wild Type as baseline)
    JAK = factor(JAK, levels = c("wt", "mut"),
                 labels = c("Wild Type (Normal)", "Mutated")),
    
  ) %>%
  select(id, Arm, Sites, LDH, TMB, JAK, pfs_time, pfs_status, os_time, os_status)


```
```{r}
names(clean_df)
```

```{r}
# Numeric summary

numeric_summary <- SECOMBIT %>%
summarise(
mean_OS    = mean(OS, na.rm = TRUE),
sd_OS      = sd(OS, na.rm = TRUE),
median_OS  = median(OS, na.rm = TRUE),
mean_PFS   = mean(`PFS TOTAL`, na.rm = TRUE),
sd_PFS     = sd(`PFS TOTAL`, na.rm = TRUE),
median_PFS = median(`PFS TOTAL`, na.rm = TRUE)
)

numeric_summary
```

```{r}
#Summary statistics for categorical variables

label(clean_df$Sites) <- "Sites"
label(clean_df$LDH)   <- "LDH"
label(clean_df$TMB)   <- "TMB"
label(clean_df$JAK)   <- "JAK"

table1(~ Sites + LDH + TMB + JAK | Arm,
       data = clean_df,
       overall = FALSE)


table_cat <- table1(~ LDH | Arm, data = clean_df)
table_cat

mortality_plot <- ggplot(clean_df, aes(x = Arm, fill = as.factor(LDH)))+
geom_bar(position = "dodge")+
labs(
title = "LDH levels by ARM",
x = "ARM",
y = "Levels",
fill = "LDH"
) +
scale_fill_manual(values = c("pink", "black", "yellow"),
labels = c("Missing","Normal", "Elevated")) +
theme_minimal()

ggplotly(mortality_plot)
```

```{r}
#ARM-wise summary

arm_summary <- clean_df %>%
group_by(Arm) %>%
summarise(
n = n(),
mean_OS = mean(os_status, na.rm = TRUE),
sd_OS = sd(os_status, na.rm = TRUE),
mean_PFS = mean(pfs_status, na.rm = TRUE),
sd_PFS = sd(pfs_status, na.rm = TRUE),
.groups = "drop"
)

arm_summary
```
```{r}
event_count <- clean_df %>%
  group_by(Arm)  %>%
  summarise(
    OS_Events = sum(os_status == 1, na.rm = TRUE),
    PFS_Events = sum(pfs_status == 1, na.rm = TRUE),
    .groups = "drop"
  )
event_count

event_long <- event_count %>%
  pivot_longer(
    cols = c(OS_Events, PFS_Events),
    names_to = "Endpoint",
    values_to = "Events"
  )

plot_events <- ggplot(event_long, aes(x = Arm, y = Events, fill = Endpoint)) +
  geom_col (position = position_dodge(width = 0.7))+
  labs(
    title = "OS and PFS : Event Vs Arms",
    x = "ARM",
    y = "No. of events",
    fill = "Endpoint",
  ) +
scale_fill_manual(values = c("OS_Events"  = "red", 
                             "PFS_Events" = "blue"))+
theme_minimal()

ggplotly(plot_events)

```
```{r}
#Baseline variables

vars <- c("Sites", "LDH", "TMB", "JAK", "os_status", "pfs_status")
factorVars <- c("Sites", "LDH", "TMB", "JAK")

tab1 <- CreateTableOne(
vars = vars,
strata = "Arm",
data = clean_df,
factorVars = factorVars
)

print(tab1, showAllLevels = TRUE, quote = FALSE, noSpaces = TRUE)
```

```{r}
par(mar = c(3, 3, 1.1, 1))
fit_os <- survfit(Surv(os_time, os_status) ~ Arm, data = clean_df)

ggsurvplot(
  fit_os, data = clean_df,
  xlab = "Time (months)", ylab = "Survival probability",
  title = "Overall Survival (OS)",
  risk.table = FALSE, pval = TRUE, conf.int = TRUE
)
```
```{r km-pfs, echo=FALSE, fig.height=2.2, fig.width=6, out.width="100%"}
par(mar = c(3, 3, 1.1, 1))
fit_pfs <- survfit(Surv(pfs_time, pfs_status) ~ Arm, data = clean_df)

ggsurvplot(
  fit_pfs, data = clean_df,
  xlab = "Time (months)", ylab = "Survival probability",
  title = "Progression-Free Survival (PFS)",
  risk.table = FALSE, pval = TRUE, conf.int = TRUE
) 
```

```{r}
#LogRank test

logrank_os  <- survdiff(Surv(os_time, os_status) ~ Arm, data = clean_df)
logrank_pfs <- survdiff(Surv(pfs_time, pfs_status) ~ Arm, data = clean_df)

p_os  <- 1 - pchisq(logrank_os$chisq,  df = length(logrank_os$n)  - 1)
p_pfs <- 1 - pchisq(logrank_pfs$chisq, df = length(logrank_pfs$n) - 1)

logrank_tab <- data.frame(
  Endpoint = c("OS", "PFS"),
  `Chi-square` = round(c(logrank_os$chisq, logrank_pfs$chisq), 2),
  df = c(length(logrank_os$n)-1, length(logrank_pfs$n)-1),
  `p-value` = signif(c(p_os, p_pfs), 3)
)

knitr::kable(logrank_tab, format = "html", align = "lccc") |>
  kableExtra::kable_styling(full_width = TRUE, font_size = 16)

```

**Inference - Survival curves and Cox Log-Rank tests**

Survival curves and log‑rank tests
The Kaplan–Meier curves for OS appear relatively close, and the log‑rank test p‑value around 0.25 suggests no statistically significant difference in overall survival across arms at this stage.

The PFS curves separate more clearly, with a log‑rank p‑value near 0.02, indicating evidence that at least one arm differs in progression‑free survival, consistent with SECOMBIT reports that sequencing can affect progression endpoints.
```{r}
#Cox proportional Hazards model


library(broom)

cox_os  <- coxph(Surv(os_time, os_status) ~ Arm + Sites + LDH + TMB + JAK,
                 data = clean_df, ties = "efron")
cox_pfs <- coxph(Surv(pfs_time, pfs_status) ~ Arm + Sites + LDH + TMB + JAK,
                 data = clean_df, ties = "efron")

tidy_cox <- function(fit, title){
  broom::tidy(fit, exponentiate = TRUE, conf.int = TRUE) |>
    mutate(
      HR_CI = sprintf("%.2f (%.2f–%.2f)", estimate, conf.low, conf.high),
      p = signif(p.value, 3)
    ) |>
    select(term, HR_CI, p) |>
    rename(Variable = term, `HR (95% CI)` = HR_CI, `p-value` = p) |>
    (\(df){
      cat(paste0("<b>", title, "</b>"))
      knitr::kable(df, format="html", escape=TRUE) |>
        kableExtra::kable_styling(full_width=TRUE, font_size=18)
    })()
}

tidy_cox(cox_os,  "Cox model — Overall Survival (OS)")
cat("<br>")
tidy_cox(cox_pfs, "Cox model — Progression-Free Survival (PFS)")


```
The multivariable Cox proportional hazards model showed no statistically significant associations between the examined covariates and survival outcome (both OS and PFS). All variables had hazard ratios with 95% confidence intervals that included 1.0 and p-values exceeding 0.05.

#COX Model for OS

In the Cox model for OS, the hazard ratios for Arm B and Arm C versus Arm A are close to 1 with wide confidence intervals and non‑significant p‑values, suggesting no clear survival advantage yet for any single sequencing strategy.

Some baseline factors, such as number of metastatic sites and LDH elevation, typically show increased mortality risk in melanoma trials, so if their HRs are above 1 with low p‑values, this aligns with prior evidence that greater tumour burden and elevated LDH are adverse prognostic markers.

#COX Model for PFS

The Cox model results for progression‑free survival show no statistically significant effects for treatment arms or most covariates, though JAK mutation and Arm B trends toward better outcomes.

JAK mutated has the strongest signal with HR 0.41 (95% CI: 0.16–1.03, p=0.057), nearly halving progression risk and just missing significance; this matches SECOMBIT biomarker findings where JAK‑mutated patients had unexpectedly good PFS and brain metastasis‑free survival.

Sites ≥3 (HR 1.60, 95% CI: 0.80–3.20, p=0.187) and elevated LDH (HR 1.14, 95% CI: 0.56–2.34, p=0.715) show increased risk trends, reflecting tumour burden as an adverse factor, though wide CIs suggest limited power.

TMB ≥10 has HR 0.85 (95% CI: 0.40–1.78, p=0.666), a neutral protective trend, in line with high TMB benefiting immunotherapy‑containing arms.

Covariates such as elevated LDH, high TMB, or multiple sites, when associated with HRs above 1, reflect faster progression, again consistent with external evidence that high tumour burden and certain biomarker profiles predict shorter progression‑free survival.

### PH Test

```{r ph-global, results="asis", echo=FALSE}
ph_os  <- cox.zph(cox_os)
ph_pfs <- cox.zph(cox_pfs)

ph_tab <- data.frame(
  Endpoint = c("OS", "PFS"),
  `PH test p-value` = signif(c(ph_os$table["GLOBAL","p"], ph_pfs$table["GLOBAL","p"]), 3)
)

knitr::kable(ph_tab, format="html") |>
  kableExtra::kable_styling(full_width=TRUE, font_size=18)
```
**Summary**

The PH test p-values indicate moderate violations of the proportional hazards assumption in the OS model and a more serious violation in the PFS model.

**OS model (p = 0.0581)**
This borderline p-value suggests the proportional hazards assumption holds approximately for overall survival, but there's weak evidence against it.

Hazard ratios from the Cox model remain reasonably interpretable, though caution is warranted; the curves may cross or diverge over time.
​

** PFS model (p = 0.0119)**

The low p-value clearly rejects proportional hazards for progression-free survival, meaning treatment effects likely change over follow-up time.

This justifies moving beyond standard Cox to time-varying effects, stratified models, or multi-state approaches.

#Win-ratio 
## B Vs A
```{r}
wr_ba <- winratio(id = "id",
                  trt = "Arm",
                  outcomes = list(
                    outc1 = c("os_status", "s", "os_time"),
                    outc2 = c("pfs_status", "s", "pfs_time")
                  ),
                  fu = "os_time",
                  data = clean_df %>% filter(Arm != "C") %>%
                    filter(!is.na(pfs_time)| !is.na(os_time)) %>%
                    mutate(pfs_time = round(pfs_time, 1),
                           os_time = round(os_time, 1))
)
summary(wr_ba)
```
* This win-ratio analysis favours ARM B over ARM A with a clincially meaningful point estimate but lacks statistical significance.

* Out of 4761 paired comparisons between Arms A and B (with 69 patients in each arm), ARM B **wins** 2289 pairs while ARM A wins 1541, with a win-ratio of 1.49 and a p-value of 0.0993.

* Wins are more driven by OS (1733 Vs 1271) than PFS (556 Vs 270). This reflects ARM B's edge in both endpoints despite individual COX models with non-significant Hazard ratios.

* This aligns with SECOMBIT's longer-term data favoring Arms B & C.


## C Vs A
```{r}

wr_ca <- winratio(id = "id",
                  trt = "Arm",
                  outcomes = list(
                    outc1 = c("os_status", "s", "os_time"),
                    outc2 = c("pfs_status", "s", "pfs_time")
                  ),
                  fu = "os_time",
                  data = clean_df %>% filter(Arm != "B") %>%
                    filter(!is.na(pfs_time)| !is.na(os_time)) %>%
                    mutate(pfs_time = round(pfs_time, 1),
                           os_time = round(os_time, 1))
)

summary(wr_ca)
```
The win-ratio analysis show Arm C has a statistically significant advantage over Arm A, with a superior OS performance.

* Out of 4692 paired comparisons between Arms C and A (with 68 patients in Arm C and 69 in Arm A), ARM C wins 2361 pairs while ARM A wins only 143, with a win-ratio of 1.58 and p-value is at 0.0508.

* Strong OS contribution dominates with PFS also favouring ARM C. 

* Few ties than other analyses sharpen the signal.

* Both Arm C and A show no benefit over control, with Arm C has nominal significance despite similar sample sizes.

## C Vs B
```{r}

wr_cb <- winratio(id = "id",
                  trt = "Arm",
                  outcomes = list(
                    outc1 = c("os_status", "s", "os_time"),
                    outc2 = c("pfs_status", "s", "pfs_time")
                  ),
                  fu = "os_time",
                  data = clean_df %>% filter(Arm != "A") %>%
                    filter(!is.na(pfs_time)| !is.na(os_time)) %>%
                    mutate(pfs_time = round(pfs_time, 1),
                           os_time = round(os_time, 1))
)

summary(wr_cb)
```
* The win-ratio analysis show no meaningful advantage for Arm C over Arm B, with a near-neutral point estimate and no statistical significance.

* Out of 4692 paired comparisons between Arms C and B (with 68 patients in Arm C and 69 in Arm B), ARM C wins 1646 pairs while ARM B wins 1562, with a win-ratio of 1.05 and p-value is at 0.8410.

* OS wins are balanced and PFS slightly favours ARM C. 

* High ties and similar totals indicates equivalence.

* Both Arm B and C outperform A without any clear superiority between them.


```{r}
#Adjusted p-values for multiple testing

pvals <- c(
  "B_vs_A" = wr_ba$p.value,
  "C_vs_B" = wr_cb$p.value,
  "C_vs_A" = wr_ca$p.value
)

p.adjust(pvals, method = "bonferroni")

```
After multiple testing adjustment using Bonferroni, none of the win-ratio comparisions between arms shows statistical significance.

* **B Vs A:** Original p=0.099 becomes 0.298 (non-significant)

* **C Vs B:** Original p=0.841 becomes 1.000 (non-significant)

* **C Vs A:** Original p=0.051 becomes 0.1525 (non-significant)

* Bonferroni correction for 3 comparisons appropriately controls arm-wise error but reduces power with approximately 68-69 patients in each arm

* Unadjusted trends - Arm B and C both favor over Arm A but lacks confirmation.

```{r}
# **NET BENEFIT**
# B Vs A

nb_df_BA <- clean_df %>%
  filter(Arm %in% c("A","B")) %>%
  filter(!is.na(os_time) | !is.na(pfs_time)) %>%   # keep if at least one endpoint present
  mutate(
    trt = factor(Arm, levels = c("A","B"))
  )

nb_BA <- BuyseTest(
  trt ~ 
    tte(os_time, os_status, threshold = 0) +   # priority 1: OS
    tte(pfs_time, pfs_status, threshold = 0),  # priority 2: PFS
  data = nb_df_BA,
  hierarchical = TRUE
)

summary(nb_BA)

```
* Significant PFS contribution (p-value 0.031)

* Global results reflects 22% net benefit for B with OS alone contributing for 13% with 8.6% from PFS. 

```{r}
# C Vs A
nb_df_CA <- clean_df %>%
  filter(Arm %in% c("A","C")) %>%
  filter(!is.na(os_time) | !is.na(pfs_time)) %>%
  mutate(trt = factor(Arm, levels = c("A","C")))   # A control, C treatment

nb_CA <- BuyseTest(
  trt ~ 
    tte(os_time, os_status, threshold = 0) +
    tte(pfs_time, pfs_status, threshold = 0),
  data = nb_df_CA,
  hierarchical = TRUE
)

summary(nb_CA)

```
* Generalised pairwise comparisons for Arm C vs Arm A is similar to Arm B Vs Arm A.

* Global net benefit :  0.2015 - Arm C beats Arm A by overall 20.15% with 13.23% contributed by the OS. 

* Strong PFS signal noted (p=0.0499)

```{r}
# C Vs B

nb_df_CB <- clean_df %>%
  filter(Arm %in% c("B","C")) %>%
  filter(!is.na(os_time) | !is.na(pfs_time)) %>%
  mutate(trt = factor(Arm, levels = c("B","C")))   # B control, C treatment

nb_CB <- BuyseTest(
  trt ~ 
    tte(os_time, os_status, threshold = 0) +
    tte(pfs_time, pfs_status, threshold = 0),
  data = nb_df_CB,
  hierarchical = TRUE
)

summary(nb_CB)

```
* Generalised pairwise comparisons confirm equivalence between Arms C and B with no net benefit.

* OS contribution -1% net confirming long term survival, PFS -3.89% is clinically not significant.


**Multi-state model**

State 1: Stable
State 2: Progression
State 3: Death

```{r}
states <- c("Stable", "Progression", "Death")

```

```{r}
#Define transition intensity matrix (Q)

Q <- matrix(0, nrow = 3, ncol = 3, dimnames = list(states, states))

Q["Stable","Progression"] <- 0.1
Q["Stable","Death"]       <- 0.05
Q["Progression","Death"]  <- 0.1

# Other transitions remain 0 (impossible)

```


```{r}

# Prepare data in msm long format

msm_df <- clean_df %>%
  select(id, Arm, os_time, os_status, pfs_time, pfs_status) %>%
  rowwise() %>%
  do({
    id <- .$id
    arm <- .$Arm
    
    # Start in Stable at time 0
    df_list <- list(
      data.frame(id = id, time = 0, state = 1, Arm = arm)
    )
    
    # Progression, only if before death
    if(!is.na(.$pfs_time) && .$pfs_status == 1 && 
       (is.na(.$os_time) || .$pfs_time < .$os_time)){
      df_list <- c(df_list, list(data.frame(id = id, time = .$pfs_time, state = 2, Arm = arm)))
    }
    
    # Death, if occurred
    if(!is.na(.$os_time) && .$os_status == 1){
      df_list <- c(df_list, list(data.frame(id = id, time = .$os_time, state = 3, Arm = arm)))
    }
    
    bind_rows(df_list)
  }) %>%
  ungroup() %>%
  arrange(id, time)


```


```{r}
#Multi-state  model

# Arm as covariate affecting all transitions
msm_fit <- msm(
  state ~ time,
  subject = id,
  data = msm_df,
  qmatrix = Q,
  covariates = ~ Arm,
  exacttimes = TRUE
)

# Summary
summary(msm_fit)
```

```{r}
# Probabilities from Stable at time t
pt <- pmatrix.msm(msm_fit, t = 12)   # e.g., 12 months
pt

# Example: plot transition probabilities from Stable over time
times <- seq(0, 24, by = 1)  # 0–24 months
probs <- sapply(times, function(t) pmatrix.msm(msm_fit, t = t)[1, ])
matplot(times, t(probs), type = "l", lty = 1, col = 1:3,
        xlab = "Time (months)", ylab = "Probability",
        main = "State occupation probabilities from Stable")
legend("topright", legend = states, col = 1:3, lty = 1)

```


```{r}
pt_A_12 <- pmatrix.msm(msm_fit, t = 12, covariates = list(Arm = "A"))
pt_B_12 <- pmatrix.msm(msm_fit, t = 12, covariates = list(Arm = "B"))
pt_C_12 <- pmatrix.msm(msm_fit, t = 12, covariates = list(Arm = "C"))

pt_A_12
pt_B_12
pt_C_12
```

```{r}
probs_A <- sapply(times, function(t)
  pmatrix.msm(msm_fit, t = t, covariates = list(Arm = "A"))[1, ])

probs_B <- sapply(times, function(t)
  pmatrix.msm(msm_fit, t = t, covariates = list(Arm = "B"))[1, ])

probs_C <- sapply(times, function(t)
  pmatrix.msm(msm_fit, t = t, covariates = list(Arm = "C"))[1, ])

```

```{r}
#Plots
#Probability of remaining Stable by Treatment
plot(times, probs_A[1,], type = "l", lwd = 2,
     ylim = c(0,1),
     xlab = "Time (months)",
     ylab = "P(Stable)",
     main = "Probability of remaining Stable by Treatment")

lines(times, probs_B[1,], lwd = 2, lty = 2)
lines(times, probs_C[1,], lwd = 2, lty = 3)

legend("topright",
       legend = c("Arm A", "Arm B", "Arm C"),
       lty = c(1,2,3), lwd = 2)


#Cumulative probability of Death by Treatment
plot(times, probs_A[3,], type = "l", lwd = 2,
     ylim = c(0,1),
     xlab = "Time (months)",
     ylab = "P(Death)",
     main = "Cumulative probability of Death by Treatment")

lines(times, probs_B[3,], lwd = 2, lty = 2)
lines(times, probs_C[3,], lwd = 2, lty = 3)

legend("topleft",
       legend = c("Arm A", "Arm B", "Arm C"),
       lty = c(1,2,3), lwd = 2)

```

```{r}
summary_df <- data.frame(
  Time = c(6, 12, 24),
  Stable_A = probs_A[1, times %in% c(6,12,24)],
  Stable_B = probs_B[1, times %in% c(6,12,24)],
  Stable_C = probs_C[1, times %in% c(6,12,24)],
  Death_A  = probs_A[3, times %in% c(6,12,24)],
  Death_B  = probs_B[3, times %in% c(6,12,24)],
  Death_C  = probs_C[3, times %in% c(6,12,24)]
)

summary_df
```